---
title: "Upper Walla Walla Watershed - Stream Temperature Assessment"
author:
  - Bryce N. Oldemeyer:
      email: bryce.oldemeyer@mthoodenvironmental.com
      institute: [mhe_challis]
      correspondence: true
  - Mark Roes:
      email: mark.roes@mthoodenvironmental.com
      institute: [mhe_sandy]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: wdfw
      correspondence: false
  - Joshua P. Egan:
      email: joshua.egan@mthoodenvironmental.com
      institute: [mhe_moscow]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [mhe_mccall]
      correspondence: true
institute:
  - mhe_challis: Mount Hood Environmental, PO Box 1303, Challis, Idaho 83226, USA
  - mhe_sandy: Mount Hood Environmental, 39085 Pioneer Boulevard \#100 Mezzanine, Sandy, Oregon, 97055, USA
  - wdfw: Washington Department of Fish and Wildife, Fish Program, Science Division, 1111 Washington Street NE, Olympia, Washington, 98501, USA
  - mhe_moscow: Mount Hood Environmental, Temporary Moscow Address, Idaho xxxxx, USA
  - mhe_mccall: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    number_sections: yes
    pandoc_args:
    - --lua-filter=../templates/scholarly-metadata.lua
    - --lua-filter=../templates/author-info-blocks.lua
    - --lua-filter=../templates/pagebreak.lua  
csl: "../templates/journal-of-archaeological-science.csl"
bibliography:
  - AckermanLibrary.bib
always_allow_html: yes
---

<!-- the following inserts MHE logo into header -->

```{=html}
<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"logo.jpg\" style=\"float: right;width: 150px;\"/>')
   });
</script>
```
```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 600
)

options(knitr.kable.NA = '--')

```

```{r load-libraries}
# run this if running chunks directly

# for formatting
library(kableExtra)
library(ggpubr)

# for analysis
library(sf)
library(here)
library(tidyverse)
library(magrittr)
library(ggmap)
library(nngeo)
library(janitor)

theme_set(theme_pubr(x.text.angle = 45,
                     base_size = 8))

```

```{r load-data}

# Project area polygons and 200m layer (with habitat attributes) within project area

load(here("analysis/data/derived_data/uww_spatial.rda"))


```

# Background

Water is the cornerstone of First Foods production and inherently tied to the five fundamental touchstones: hydrology, geomorphology, habitat and network connectivity, riverine biotic community, and riparian vegetation [@Jones2008]. These touchstones form the Umatilla River Vision, a holistic approach to water quality restoration. Add description of stream assessment

## Study Area

This assessment will be focused on the alluvial channel of the Walla Walla River from the confluence with Dry Creek near Lowden, Washington, to the headwaters of the North and South forks of the Walla Walla River. The primary study area includes approximately 70 miles of stream. The secondary study area includes the catchment of the primary study area which encompasses approximately 885 square miles.

# Stream Temperature Assessment

The primary objectives of the Upper Walla Walla stream temperature assessment is to:

-   Summarize current stream temperatures for the Upper Walla Walla using NorWest stream temperature data (CITE)

-   Compare predictions of stream temperatures in the Upper Walla Walla for 2040 and 2080 climate scenarios.

-   Summarize temperature suitability based solely on species-specific temperature models for the Upper Walla Walla.

-   Run a hypothetical scenario were restoration efforts over the next \~60 years help decrease stream temperatures by -1^o^ C

## Stream Temperature Data

NorWest August stream temperature estimates -- 2002-2011, 2040, and 2080 summaries. Additionally, -1^o^ C temperature scenario.

-   S2_02_11 Historical composite scenario representing 10 year average August mean stream temperatures for 2002-2011

-   S30_2040D Future August mean stream temperature scenario based on global climate model ensemble average projected changes in August air temperature and stream discharge for the A1B warming trajectory in the 2040s (2030-2059). Future stream deltas within a NorWeST unit account for differential sensitivity among streams so that cold streams warm less than warm streams.

-   S32_2080D Future August mean stream temperature scenario based on global climate model ensemble average projected changes in August air temperature and stream discharge for the A1B warming trajectory in the 2080s (2070-2099). Future stream deltas within a NorWeST unit account for differential sensitivity among streams so that cold streams warm less than warm streams.

## Fish Temperature Thresholds

The focal species for the stream temeprature suitability models are:

-   Spring-run Chinook salmon (*Oncorhynchus tshawytscha*; hereafter Chinook salmon)
-   Middle Columbia River summer steelhead (*O. mykiss*; hereafter steelhead; ESA-listed threatened)
-   Columbia River bull trout (*Salvelinus confluentus*; ESA-listed threatened)
-   Pacific lamprey (*Lampetra tridentata*)

# Results

## Stream Temperature Results

```{r current_temp_map}

river_color = "lightskyblue1"

# Current temp map
temp_map_cur = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA)),
          aes(color = S2_02_11),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2002-2011 historical composite scenario",
       color = expression("Temperature ("^o*"C)"))

# 2040 temp map
temp_map_2040 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S30_2040D = replace(S30_2040D, S30_2040D == -9999.00, NA)),
          aes(color = S30_2040D),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2040 future scenario",
       color = expression("Temperature ("^o*"C)"))

# 2080 temp map
temp_map_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA)),
          aes(color = S32_2080D),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2080 future scenario",
       color = expression("Temperature ("^o*"C)"))

# -1C results
temp_map_1C_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA)),
          aes(color = (S32_2080D-1)),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = expression("Hypothetical restoration effect on \n 2080 future scenario (-1"^o*"C)"),
       color = expression("Temperature ("^o*"C)"))


# 2040 comparison
temp_rel_map_2040 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S30_2040D = replace(S30_2040D, S30_2040D == -9999.00, NA),
                   S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
                   temp_dif = S30_2040D - S2_02_11),
          aes(color = temp_dif),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(0,2.5)) +
#scale_fill_gradientn(limits = c(1,3)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2040 future scenario",
       color = expression("Temperature ("^o*"C)"))

# 2080 comparison
temp_rel_map_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA),
                   S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
                   temp_dif = S32_2080D - S2_02_11),
          aes(color = temp_dif),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(0,2.5)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2080 future scenario",
       color = expression("Temperature ("^o*"C)"))


# -1C results
temp_rel_map_1C_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA),
                   S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
                   temp_dif = S32_2080D - S2_02_11),
          aes(color = (temp_dif-1)),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(0,2.5)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = expression("Hypothetical restoration effect on \n 2080 future scenario (-1"^o*"C)"),
       color = expression("Temperature ("^o*"C)"))

```

```{r current-temp-maps, fig.cap = "NorWeST average August mean stream temperatures for the Upper Walla Walla."}

ggarrange(temp_map_cur,temp_map_2040,temp_map_2080,temp_map_1C_2080)

```

```{r temp-comparison-maps, fig.cap = "Comparison of future NorWeST average August mean stream temperatures for  2040, 2080, and 2080 - 1oC to the 2002-2011 historical composite scanerio for the Upper Walla Walla."}

ggarrange(temp_map_cur,temp_rel_map_2040,temp_rel_map_2080, temp_rel_map_1C_2080)

```

## Fish Temperature Suitability Results

Chinook map & summary table

Steelhead map & summary table

Bull trout map & summary table

Lamprey map & summary table

# Stream Temperature Discussion
