---
title: "Summer Stream Temperature Assessment, Upper Walla Walla Watershed"
author:
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [mhe_mccall]
      correspondence: true
  - Bryce N. Oldemeyer:
      email: bryce.oldemeyer@mthoodenvironmental.com
      institute: [mhe_challis]
      correspondence: true
  - Mark Roes:
      email: mark.roes@mthoodenvironmental.com
      institute: [mhe_sandy]
      correspondence: false
institute:
  - mhe_mccall: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
  - mhe_challis: Mount Hood Environmental, PO Box 1303, Challis, Idaho 83226, USA
  - mhe_sandy: Mount Hood Environmental, 39085 Pioneer Boulevard \#100 Mezzanine, Sandy, Oregon, 97055, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    number_sections: yes
    pandoc_args:
    - --lua-filter=../templates/scholarly-metadata.lua
    - --lua-filter=../templates/author-info-blocks.lua
    - --lua-filter=../templates/pagebreak.lua  
csl: "../templates/journal-of-archaeological-science.csl"
bibliography:
  - AckermanLibrary.bib
always_allow_html: yes
---

<!-- the following inserts MHE logo into header -->

```{=html}
<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"logo.jpg\" style=\"float: right;width: 150px;\"/>')
   });
</script>
```

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 600
)

options(knitr.kable.NA = '--')

```

```{r load-libraries}
# run this if running chunks directly

# for formatting
library(kableExtra)
library(ggpubr)

# for analysis
library(sf)
library(here)
library(tidyverse)
library(magrittr)
library(ggmap)
library(nngeo)
library(janitor)

theme_set(theme_pubr(x.text.angle = 45,
                     base_size = 8))

```

```{r load-data}
# Project area polygons and 200m layer (with habitat attributes) within project area
load(here("analysis/data/derived_data/uww_spatial.rda"))

```


# Background

Water is the cornerstone of First Foods production and inherently tied to the five fundamental touchstones: hydrology, geomorphology, habitat and network connectivity, riverine biotic community, and riparian vegetation [@Jones2008]. These touchstones form the Umatilla River Vision, a holistic approach to water quality restoration. Stream temperature is a critical component of the hydrology of the Upper Walla Walla watershed, directly affecting the fish community and the connectivity of their habitat. Warming climatic conditions are expected to have pronounced effects on salmonids [@Kovach2016]. In the Pacific Northwest, summer months are anticipated to become progressively more stressful as stream temperatures increase with warming air temperatures, which is likely to shift and reduce suitable habitat for many species [@Isaak2015; @Jones2014]. Here, we evaluate recent composite and future projected summer stream temperature scenarios [@Isaak2017a] for the Upper Walla Walla watershed and compare them to life-stage specific temperature threshholds for the target species. The goal of this assessment is to determine whether summer stream temperature is a potential factor limiting fish population productivity and to identify if and where restoration actions (e.g., riparian planting, improved grazing practices) aimed at reducing stream temperatures should be considered.


## Objectives

The primary objectives of the Upper Walla Walla summer stream temperature assessment include:

1. Summarize recent composite and future projected summer stream temperature scenarios for the Upper Walla Walla available from the [NorWeST](https://www.fs.usda.gov/rm/boise/AWAE/projects/NorWeST.html) database [Isaak2017a].

2. Compare current and projected summer stream temperature scenarios in the Upper Walla Walla to life-stage specific temperature thresholds for target species to determine whether temperatures may hinder summer life-stages of those species.

3. Evaluate a hypothetical scenario where restoration efforts (e.g., riparian planting, improved grazing practices) over the next \~60 years decrease summer stream temperatures relative to current temperatures by -1^o^C.


# Methods

We used recent historic (2002-2011) and projected (2040 & 2080) August stream temperature predictions [@Isaak2017a] for the Upper Walla Walla River watershed plus life-stage specific temperature thresholds available from the literature for target species (Table \@ref(tab:temp-thresh)) to evaluate whether current or future projected summer stream temperatures might hinder the ability of the species to use available habitat during summer months. To accomplish this, we calculated whether recent composite and projected August mean stream temperatures exceeded optimum or suitable temperature thresholds for life-stages present during summer months. Finally, we evaluated a hypothetical scenario where summer stream temperatures were reduced by 1^o^C relative to recent composite temperatures. 


## Study Area

The summer stream temperature assessment is focused on the upper Walla Walla River from the confluence with Dry Creek near Lowden, Washington, to the headwaters of the North and South forks of the Walla Walla River and include streams and tributaries (Table \@ref(tab:study-area)) modeled as part of the NorWeST regional database [@Isaak2017a].

```{r study-area}
study_tab = uww_rch_sf %>%
  select(GNIS_Name, reach_leng, S2_02_11) %>%
  st_drop_geometry() %>%
  filter(!is.na(S2_02_11)) %>%
  group_by(GNIS_Name) %>%
  summarize(`Length (km)` = round(sum(reach_leng) / 1000, 1)) %>%
  rename(`Stream Name` = GNIS_Name) %>%
  adorn_totals() %>%
  mutate(`Stream Name` = ifelse(is.na(`Stream Name`), "Unnamed Creek(s)", `Stream Name`)) %>%
  kable(booktabs = T,
        align = "lc",
        caption = "Streams included in the summer stream temperature assessment that are modeled as part of the NorWeST regional database.") %>%
  kable_styling(full_width = F,
                position = "center",
                bootstrap_options = c("striped", "condensed"))
study_tab

```


## Target Species

The focal species for the summer stream temperature assessment include:

* Spring-run Chinook salmon (*Oncorhynchus tshawytscha*; hereafter Chinook salmon)
* Middle Columbia River summer steelhead (*O. mykiss*; hereafter steelhead; ESA-listed threatened)
* Columbia River bull trout (*Salvelinus confluentus*; ESA-listed threatened)
* Pacific lamprey (*Lampetra tridentata*)


## Summer Stream Temperature Scenarios

Three NorWeST summer stream tempature scenarios were considered:

* **Recent Mean** (S2_02_11): A historical composite scenario representing the 10-year average August mean stream temperatures for 2002-2011.

* **2040 Projected** (S30_2040D): A future August mean stream temperature scenario based on global ensemble average projected changes in August air temperature and stream discharge for A1B warming trajectory in the 2040s (2030-2059). Future stream deltas within a NorWeST unit account for differential sensitivity among streams so that cold stream warm less than warm streams.

* **2080 Projected** (S32_2080D): A future August mean stream temperature scenario based on global ensemble average project changes in August air temperature and stream discharge for A1B warming trajectory in the 2080s (2070-2099). Future stream deltas within a NorWeST unit account for differential sensitivity among streams so that cold streams warm less than warm streams.

```{r norwest, fig.cap = "Recent mean (2002-2011) and projected (2040 and 2080) August mean stream temperature scenarios available from NorWeST (Isaak et al. 2017) for the Upper Walla Walla watershed."}
river_color = "lightskyblue1"
norwest_df = uww_rch_sf %>%
  select(S2_02_11,
         S30_2040D,
         S32_2080D,
         reach_leng) %>%
  mutate(S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
         S30_2040D = replace(S30_2040D, S30_2040D == -9999.00, NA),
         S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA)) %>%
  rename(`Recent Mean` = S2_02_11,
         `2040 Projected` = S30_2040D,
         `2080 Projected` = S32_2080D) %>%
  pivot_longer(cols = c("Recent Mean",
                        "2040 Projected",
                        "2080 Projected"),
               names_to = "scenario",
               values_to = "temp_c") %>%
  mutate(scenario = factor(scenario, levels = c("Recent Mean", 
                                                "2040 Projected", 
                                                "2080 Projected")))

norwest_p = norwest_df %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = norwest_df,
          aes(color = temp_c),
          size = 1) +
  scale_color_viridis_c(option = "turbo") +
  facet_wrap(~scenario, 
             ncol = 2) +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "NorWeST Summer Stream Temperatures",
       color = expression("Temperature ("^o*"C)"))
norwest_p

```

```{r norwest-bp, out.width = "70%", fig.cap = "Boxplots showing distributions of recent mean (2002-2011) and projected (2040 and 2080) August mean stream temperatures from NorWeST (Isaak et al. 2017) for the Upper Walla Walla watershed."}
norwest_bp = norwest_df %>%
  ggplot() +
  geom_boxplot(aes(x = reorder(scenario, temp_c),
                   y = temp_c),
               fill = "coral1") +
  theme_bw() +
  labs(y = expression("Temperature ("^o*"C)")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(size = 12,
                                   color = "black"),
        axis.text.y = element_text(size = 12,
                                   color = "black"))
norwest_bp

```


## Fish Temperature Threshholds

For Chinook salmon, we considered the spawning and summer parr life-stages, and for steelhead, the summer parr life-stage was considered (Table \@ref(tab:temp-thresh)). Temperature thresholds were from a literature review conducted by @Carter2005 and are expressed as the maximum weekly maximum temperature (MWMT). Chinook spawning has been observed in a wide range of temperatures (5.6 - 17.7^o^C); however, a majority of temperature observations cite maximum temperatures below 14.5^o^C. For juvenile Chinook salmon rearing, 10-16^o^C represents an optimal range for growth and rearing. Temperatures above 19^o^C generally do not support Chinook salmon parr growth and temperatures above 22^o^C typically limits their presence. Juvenile steelhead temperature tolerances are similar to those of Chinook salmon, but instead have a higher end to their optimal range of 18^o^C for growth and rearing.

Bull trout tend to have more specific habitat requirements compared to other salmonids, typically needing colder stream temperatures. Although bull trout have been reported in stream temperatures above 20^o^C, they are seldom found in waters where temperatures exceed 15-16^o^C. Temperatures above 16^o^C are detrimental for bull trout feeding and growth and are unlikely to be suitable for their long-term survival [@Selong2001 and references therein]. At temperatures greater than 18^o^C, bull trout exhibit reduced consumption, growth, and survival [@Selong2001]. Pacific lamprey can tolerate temperature range of 5-25^o^C [@Clemens2017]. Pacific lamprey spawning typically occurs at 10-18^o^C with early development from about 14-19^o^C and preferred temperatures by adults of 16-17^o^C [@Clemens2016; @Clemens2017]. Temperatures of 20^o^C and higher are typically associated with stress, tissue damage, and potential mortality in Pacific lamprey [@Clemens2016].

```{r temp-thresh}
temp_thresh_tab = tibble(Species = c("Chinook", "Chinook", "Steelhead", "Bull Trout", "Pacific Lamprey"),
                         `Life Stage` = c("Spawning", "Summer Parr", "Summer Parr", "All", "All"),
                         Source = c("Carter 2005", "Carter 2005", "Carter 2005", "Selong et al. 2001", "Clemens et al. 2016, 2017"),
                         Timing = c("8/10 - 9/30", "6/10 - 10/15", "6/10 - 10/15", "7/1 - 8/31", "7/1 - 8/31"),
                         Minimum = c("NA", "4.5", "4.5", "0", "5"),
                         Optimum = c("5.6 - 14.5", "10 - 16", "10 - 18", "10.2 - 16", "10 - 17"),
                         Maximum = c("17.7", "19", "19", "18", "20"),
                         Acute = c("20", "22", "22", "20", "25")) %>%
  kable(booktabs = T,
        align = "cccccccc",
        caption = "Life stage timing, and minimum, optimum, maximum, and acute temperature (Celcius) thresholds for summer life-stages of Chinook salmon, steelhead, bull trout, and Pacific lamprey. Chinook salmon and steelhead temperatures are expressed as the 7-day Maximum Weekly Maximum Temperature (MWMT).") %>%
  kable_styling(full_width = F,
                position = "center",
                bootstrap_options = c("striped", "condensed"))
temp_thresh_tab

```


# Results

## Stream Temperature Suitability

```{r temp-suit-tab}
#temp_suit_tab = norwest_df

```

```{r optimum-map, fig.width = 6, fig.height = 7, fig.cap = "Maps depicting whether recent mean (2002-2011) and projected (2040 and 2080) August mean stream temperatures exceed (TRUE) or are below (FALSE) the top optimal temperature for summer life-stages of target species within the Upper Walla Walla watershed."}
optimum_df = norwest_df %>%
  mutate(`Chinook Spawning` = ifelse(temp_c > 14.5, T, F),
         `Chinook Parr` = ifelse(temp_c > 16, T, F),
         `Steelhead Parr` = ifelse(temp_c > 18, T, F),
         `Bull Trout` = ifelse(temp_c > 16, T, F),
         `Pacific Lamprey` = ifelse(temp_c > 17, T, F)) %>%
  pivot_longer(cols = c("Chinook Spawning",
                        "Chinook Parr",
                        "Steelhead Parr",
                        "Bull Trout",
                        "Pacific Lamprey"),
               names_to = "spec_ls",
               values_to = "above_optimum") %>%
  mutate(above_optimum = replace_na(above_optimum, F)) %>%
  mutate(spec_ls = factor(spec_ls, levels = c("Chinook Spawning",
                                              "Chinook Parr",
                                              "Steelhead Parr",
                                              "Bull Trout",
                                              "Pacific Lamprey")))

optimum_p = optimum_df %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union(),
          fill = NA,
          color = "gray50",
          size = 0.5) +
  #geom_sf(color = river_color) +
  geom_sf(data = optimum_df,
          aes(color = above_optimum),
          size = 0.5) +
  scale_color_manual(values = c("FALSE" = river_color,
                                "TRUE" = "orangered")) +
  facet_grid(spec_ls ~ scenario) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(color = "Above Optimum Range")
optimum_p

```

```{r max-map, fig.width = 6, fig.height = 7, fig.cap = "Maps depicting whether recent mean (2002-2011) and projected (2040 and 2080) August mean stream temperatures exceed (TRUE) or are below (FALSE) the maximum temperature for summer life-stages of target species within the Upper Walla Walla watershed."}
maximum_df = norwest_df %>%
  mutate(`Chinook Spawning` = ifelse(temp_c > 17.7, T, F),
         `Chinook Parr` = ifelse(temp_c > 19, T, F),
         `Steelhead Parr` = ifelse(temp_c > 19, T, F),
         `Bull Trout` = ifelse(temp_c > 18, T, F),
         `Pacific Lamprey` = ifelse(temp_c > 20, T, F)) %>%
  pivot_longer(cols = c("Chinook Spawning",
                        "Chinook Parr",
                        "Steelhead Parr",
                        "Bull Trout",
                        "Pacific Lamprey"),
               names_to = "spec_ls",
               values_to = "above_maximum") %>%
  mutate(above_maximum = replace_na(above_maximum, F)) %>%
  mutate(spec_ls = factor(spec_ls, levels = c("Chinook Spawning",
                                              "Chinook Parr",
                                              "Steelhead Parr",
                                              "Bull Trout",
                                              "Pacific Lamprey")))

maximum_p = maximum_df %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union(),
          fill = NA,
          color = "gray50",
          size = 0.5) +
  #geom_sf(color = river_color) +
  geom_sf(data = maximum_df,
          aes(color = above_maximum),
          size = 0.5) +
  scale_color_manual(values = c("FALSE" = river_color,
                                "TRUE" = "red")) +
  facet_grid(spec_ls ~ scenario) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(color = "Above Maximum Temperature")
maximum_p

```


## Habitat Capacity Results

Describe changes in QRF relative habitat capacities using the 2040, 2080, and -1 scenarios using the RF extrapolation model. Do we see decreases in capacity under projected climate change scenarios? Alternatively, could some habitat capacity limitations be rectified through rehabilitation actions to cool stream temperatures by 1C.
<!--
* Re-run the RF extrapolations for the UWW also using the 2040, 2080, and -1 scenarios.
* Compare relative changes in capacity for those scenarios relative to the Recent Mean scenario
-->


# Discussion

Discussion text...


# Literature Cited

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>


<!-- misc deprecated code and figures 
```{r current_temp_map}

river_color = "lightskyblue1"

# Current temp map
temp_map_cur = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA)),
          aes(color = S2_02_11),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2002-2011 historical composite scenario",
       color = expression("Temperature ("^o*"C)"))

# 2040 temp map
temp_map_2040 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S30_2040D = replace(S30_2040D, S30_2040D == -9999.00, NA)),
          aes(color = S30_2040D),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2040 future scenario",
       color = expression("Temperature ("^o*"C)"))

# 2080 temp map
temp_map_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA)),
          aes(color = S32_2080D),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2080 future scenario",
       color = expression("Temperature ("^o*"C)"))

# -1C results
temp_map_1C = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA)),
          aes(color = (S2_02_11-1)),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = expression("Hypothetical restoration effect on \n historic composite scenario (-1"^o*"C)"),
       color = expression("Temperature ("^o*"C)"))


# 2040 comparison
temp_rel_map_2040 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S30_2040D = replace(S30_2040D, S30_2040D == -9999.00, NA),
                   S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
                   temp_dif = S30_2040D - S2_02_11),
          aes(color = temp_dif),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(-1,2.5)) +
#scale_fill_gradientn(limits = c(1,3)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2040 future scenario",
       color = expression("Temperature ("^o*"C)"))

# 2080 comparison
temp_rel_map_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA),
                   S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
                   temp_dif = S32_2080D - S2_02_11),
          aes(color = temp_dif),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(-1,2.5)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2080 future scenario",
       color = expression("Temperature ("^o*"C)"))


# -1C results
temp_rel_map_1C = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA)),
          aes(color = (S2_02_11-1)),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(-1,2.5)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = expression("Hypothetical restoration effect on \n historic composite scenario (-1"^o*"C)"),
       color = expression("Temperature ("^o*"C)"))

```

```{r current-temp-maps, fig.cap = "NorWeST average August mean stream temperatures for the Upper Walla Walla."}

temp_map_cur

```

```{r 2040-temp-maps, fig.cap = "NorWeST average August mean stream temperatures for the Upper Walla Walla."}

temp_map_2040

```

```{r 2080-temp-maps, fig.cap = "NorWeST average August mean stream temperatures for the Upper Walla Walla."}

temp_map_2080

```

```{r 1C-temp-maps, fig.cap = "NorWeST average August mean stream temperatures for the Upper Walla Walla."}
temp_map_1C

```

```{r 2040-temp-comparison-maps, fig.cap = "Difference in projected NorWeST average August mean stream temperatures for 2040relative to the 2002-2011 historical composite scanerio for the Upper Walla Walla."}

ggarrange(temp_rel_map_2040)

```

```{r 2080-temp-comparison-maps, fig.cap = "Difference in projected NorWeST average August mean stream temperatures for 2080 relative to the 2002-2011 historical composite scanerio for the Upper Walla Walla."}

ggarrange(temp_rel_map_2080)

```

```{r 1C-temp-comparison-maps, fig.cap = "Difference in projected NorWeST average August mean stream temperatures for - 1oC scenario relative to the 2002-2011 historical composite scanerio for the Upper Walla Walla."}

ggarrange(temp_rel_map_1C)

```
-->
