---
title: "Summer Stream Temperature Assessment, Upper Walla Walla Watershed"
author:
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [mhe_mccall]
      correspondence: true
  - Bryce N. Oldemeyer:
      email: bryce.oldemeyer@mthoodenvironmental.com
      institute: [mhe_challis]
      correspondence: true
  - Mark Roes:
      email: mark.roes@mthoodenvironmental.com
      institute: [mhe_sandy]
      correspondence: false
institute:
  - mhe_mccall: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
  - mhe_challis: Mount Hood Environmental, PO Box 1303, Challis, Idaho 83226, USA
  - mhe_sandy: Mount Hood Environmental, 39085 Pioneer Boulevard \#100 Mezzanine, Sandy, Oregon, 97055, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    number_sections: yes
    pandoc_args:
    - --lua-filter=../templates/scholarly-metadata.lua
    - --lua-filter=../templates/author-info-blocks.lua
    - --lua-filter=../templates/pagebreak.lua  
csl: "../templates/journal-of-archaeological-science.csl"
bibliography:
  - AckermanLibrary.bib
always_allow_html: yes
---

<!-- the following inserts MHE logo into header -->

```{=html}
<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"logo.jpg\" style=\"float: right;width: 150px;\"/>')
   });
</script>
```

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 600
)

options(knitr.kable.NA = '--')

```

```{r load-libraries}
# run this if running chunks directly

# for formatting
library(kableExtra)
library(ggpubr)

# for analysis
library(sf)
library(here)
library(tidyverse)
library(magrittr)
library(ggmap)
library(nngeo)
library(janitor)

theme_set(theme_pubr(x.text.angle = 45,
                     base_size = 8))

```

```{r load-data}
# Project area polygons and 200m layer (with habitat attributes) within project area
load(here("analysis/data/derived_data/uww_spatial.rda"))

```

<!--
# Background

Water is the cornerstone of First Foods production and inherently tied to the five fundamental touchstones: hydrology, geomorphology, habitat and network connectivity, riverine biotic community, and riparian vegetation [@Jones2008]. These touchstones form the Umatilla River Vision, a holistic approach to water quality restoration. Add description of stream assessment

## Study Area

This assessment will be focused on the alluvial channel of the Walla Walla River from the confluence with Dry Creek near Lowden, Washington, to the headwaters of the North and South forks of the Walla Walla River. The primary study area includes approximately 70 miles of stream. The secondary study area includes the catchment of the primary study area which encompasses approximately 885 square miles.
-->

# Background and Objectives

The primary objectives of the Upper Walla Walla summer stream temperature assessment is to:

-   Summarize historic composite scenario summer stream temperatures for the Upper Walla Walla using NorWest stream temperature data (CITE)

-   Compare predictions of summer stream temperatures in the Upper Walla Walla for 2040 and 2080 climate scenarios.

-   Run a hypothetical scenario were restoration efforts over the next \~60 years decrease summer stream temperatures by -1^o^ C.

-   Summarize temperature suitability based solely on species-specific temperature models for the Upper Walla Walla.


# Approach

We used a combination of modeled NorWeST summer stream temperature predictions [@Isaak2017a] for the Upper Walla Walla River watershed plus life-stage specific temperature thresholds (Table \@ref(tab:temp-thresh)) [@Carter2005] to evaluate whether current of future projected stream temperatures might hinder the ability of Chinook salmon and steelhead to use available habitat during summer months. Three NorWeST temperature scenarios were considered:

* **S2_02_11**: A historical composite scenario representing the 10-year average August mean stream temperatures for 2002-2011.

* **S30_2040D**: A future August mean stream temperature scenario based on global ensemble average projected changes in August air temperature and stream discharge for A1B warming trajectory in the 2040s (2030-2059). Future stream deltas within a NorWeST unit account for differential sensitivity among streams so that cold stream warm less than warm streams.

* **S32_2080D**: A future August mean stream temperature scenario based on global ensemble average project changes in August air temperature and stream discharge for A1B warming trajectory in the 2080s (2070-2099). Future stream deltas within a NorWeST unit account for differential sensitivity among streams so that cold streams warm less than warm streams.

```{r norwest, fig.cap = "Current (S2_02_11) and projected (S30_2040D and S32_2080D) August mean stream temperature scenarios available from NorWeST (Isaak et al. 2017) for the Upper Walla Walla watershed."}
river_color = "lightskyblue1"
norwest_df = uww_rch_sf %>%
  select(S2_02_11,
         S30_2040D,
         S32_2080D) %>%
  mutate(S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
         S30_2040D = replace(S30_2040D, S30_2040D == -9999.00, NA),
         S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA)) %>%
  pivot_longer(cols = c("S2_02_11",
                        "S30_2040D",
                        "S32_2080D"),
               names_to = "scenario",
               values_to = "temp_c")

norwest_p = norwest_df %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = norwest_df,
          aes(color = temp_c),
          size = 1) +
  scale_color_viridis_c(option = "turbo") +
  facet_wrap(~ scenario,
             ncol = 3) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "NorWeST Stream Temperature Predictions",
       color = expression("Temperature ("^o*"C)"))
norwest_p

```

```{r temp-thresh}
temp_thresh_tab = tibble(Species = c("Chinook", "Chinook", "Steelhead"),
                         `Life Stage` = c("Spawning", "Summer Rearing", "Summer Rearing"),
                         Timing = c("8/10 - 9/30", "6/10 - 10/15", "6/10 - 10/15"),
                         Minimum = c("3.3", "4.5", "4.5"),
                         Optimum = c("7.2 - 14.5", "10 - 16", "10 - 18"),
                         Maximum = c("18", "20", "19"),
                         Acute = c("20", "22", "22")) %>%
  kable(booktabs = T,
        align = "ccccccc",
        caption = "Life stage timing, and minimum, optimum, maximum, and acute temperature (Celcius) threshholds for summer life-stages of Chinook salmon and steelhead, adopted from Carter (2005). All temperatures are expressed as the 7-day Maximum Weekly Maximum Temperature (MWMT).") %>%
  kable_styling(full_width = F,
                position = "center",
                bootstrap_options = c("striped", "condensed"))
temp_thresh_tab

```

We calculated whether current and projected August mean stream temperatures exceeded the optimum range (Table \@ref(tab:temp-thresh)) for summer life-stages of Chinook salmon and steelhead. For Chinook salmon, we considered the spawning and summer parr life-stages, and for steelhead, the summer parr life-stage was considered (Figure \@ref(fig:thresh-map)).

```{r thresh-map, out.width = "100%", fig.cap = "Maps depicting whether current (S2_02_11) and projected (S30_2040D and S32_2080D) August mean stream temperatures exceed (TRUE) or are below (FALSE) the optimal range for summer life-stages of Chinook salmon and steelhead and within the Upper Walla Walla watershed."}
thresh_df = norwest_df %>%
  mutate(`Chinook Spawning` = ifelse(temp_c > 14.5, T, F),
         `Chinook Rearing` = ifelse(temp_c > 16, T, F),
         `Steelhead Rearing` = ifelse(temp_c > 18, T, F)) %>%
  pivot_longer(cols = c("Chinook Spawning",
                        "Chinook Rearing",
                        "Steelhead Rearing"),
               names_to = "spec_ls",
               values_to = "above_optimum") %>%
  mutate(above_optimum = replace_na(above_optimum, F)) %>%
  # get reach length
  st_join(uww_rch_sf %>%
            select(geometry,
                   reach_leng))

thresh_p = thresh_df %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union(),
          fill = NA,
          color = "gray50") +
  #geom_sf(color = river_color) +
  geom_sf(data = thresh_df,
          aes(color = above_optimum),
          size = 1) +
  scale_color_manual(values = c("FALSE" = river_color,
                                "TRUE" = "red")) +
  facet_grid(spec_ls ~ scenario) +
  theme_classic() +
  theme(legend.position = "top",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(color = "Above Optimum Temperature")
thresh_p

```


## Stream Temperature Data

NorWest August stream temperature estimates for historic composite scenario (2002-2011), projected 2040 stream temperatures, and projected 2080 stream temperatures summaries.

-   S2_02_11 Historical composite scenario representing 10 year average August mean stream temperatures for 2002-2011

-   S30_2040D Future August mean stream temperature scenario based on global climate model ensemble average projected changes in August air temperature and stream discharge for the A1B warming trajectory in the 2040s (2030-2059). Future stream deltas within a NorWeST unit account for differential sensitivity among streams so that cold streams warm less than warm streams.

-   S32_2080D Future August mean stream temperature scenario based on global climate model ensemble average projected changes in August air temperature and stream discharge for the A1B warming trajectory in the 2080s (2070-2099). Future stream deltas within a NorWeST unit account for differential sensitivity among streams so that cold streams warm less than warm streams.

## Fish Temperature Thresholds

The focal species for the summer stream temperature suitability models are:

-   Spring-run Chinook salmon (*Oncorhynchus tshawytscha*; hereafter Chinook salmon)
    -   thresholds - Carter (2005) - spawning and summer parr
-   Middle Columbia River summer steelhead (*O. mykiss*; hereafter steelhead; ESA-listed threatened)
    -   thresholds - Carter (2005) - summer parr
-   Columbia River bull trout (*Salvelinus confluentus*; ESA-listed threatened)
    -   thresholds - need to find
-   Pacific lamprey (*Lampetra tridentata*)
    -   thresholds - need to find

*NOTE: Possibly adjust spatial scale to match Rio reaches, StreamNet, CTUIR? Also, export these shapefiles for additional mapping and including in the report elsewhere.*

# Results

## Stream Temperature Results

```{r current_temp_map}

river_color = "lightskyblue1"

# Current temp map
temp_map_cur = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA)),
          aes(color = S2_02_11),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2002-2011 historical composite scenario",
       color = expression("Temperature ("^o*"C)"))

# 2040 temp map
temp_map_2040 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S30_2040D = replace(S30_2040D, S30_2040D == -9999.00, NA)),
          aes(color = S30_2040D),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2040 future scenario",
       color = expression("Temperature ("^o*"C)"))

# 2080 temp map
temp_map_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA)),
          aes(color = S32_2080D),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2080 future scenario",
       color = expression("Temperature ("^o*"C)"))

# -1C results
temp_map_1C = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA)),
          aes(color = (S2_02_11-1)),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = expression("Hypothetical restoration effect on \n historic composite scenario (-1"^o*"C)"),
       color = expression("Temperature ("^o*"C)"))


# 2040 comparison
temp_rel_map_2040 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S30_2040D = replace(S30_2040D, S30_2040D == -9999.00, NA),
                   S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
                   temp_dif = S30_2040D - S2_02_11),
          aes(color = temp_dif),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(-1,2.5)) +
#scale_fill_gradientn(limits = c(1,3)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2040 future scenario",
       color = expression("Temperature ("^o*"C)"))

# 2080 comparison
temp_rel_map_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA),
                   S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
                   temp_dif = S32_2080D - S2_02_11),
          aes(color = temp_dif),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(-1,2.5)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2080 future scenario",
       color = expression("Temperature ("^o*"C)"))


# -1C results
temp_rel_map_1C = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA)),
          aes(color = (S2_02_11-1)),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(-1,2.5)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = expression("Hypothetical restoration effect on \n historic composite scenario (-1"^o*"C)"),
       color = expression("Temperature ("^o*"C)"))

```

```{r current-temp-maps, fig.cap = "NorWeST average August mean stream temperatures for the Upper Walla Walla."}

temp_map_cur

```

```{r 2040-temp-maps, fig.cap = "NorWeST average August mean stream temperatures for the Upper Walla Walla."}

temp_map_2040

```

```{r 2080-temp-maps, fig.cap = "NorWeST average August mean stream temperatures for the Upper Walla Walla."}

temp_map_2080

```

```{r 1C-temp-maps, fig.cap = "NorWeST average August mean stream temperatures for the Upper Walla Walla."}
temp_map_1C

```

```{r 2040-temp-comparison-maps, fig.cap = "Difference in projected NorWeST average August mean stream temperatures for 2040relative to the 2002-2011 historical composite scanerio for the Upper Walla Walla."}

ggarrange(temp_rel_map_2040)

```

```{r 2080-temp-comparison-maps, fig.cap = "Difference in projected NorWeST average August mean stream temperatures for 2080 relative to the 2002-2011 historical composite scanerio for the Upper Walla Walla."}

ggarrange(temp_rel_map_2080)

```

```{r 1C-temp-comparison-maps, fig.cap = "Difference in projected NorWeST average August mean stream temperatures for - 1oC scenario relative to the 2002-2011 historical composite scanerio for the Upper Walla Walla."}

ggarrange(temp_rel_map_1C)

```

## Fish Temperature Suitability Results

-   Chinook map, summary table, violin plots

    -   redds and summer parr

-   Steelhead map, summary table, violin plots

    -   summer parr

-   Bull trout map, summary table, violin plots

    -   general temp thresholds

-   Lamprey map, summary table, violin plots

    -   general temp thresholds

    *Would include maps similar to those included in Appendix A of the IRA that show the proportion of time that a water temperature was "within" a given life-stage specific threshold.*

    *Also across the 4 temperature scenarios*

    *We should think about how to summarize this in tabular form or as figures (box plots, bar plots?) form as well.*

## RF capacity estimates

-   Re-run RF capacity estimates with scenarios 2, 3, and 4

-   Chinook map, summary table, violin plots

    -   redds and summer parr

-   Steelhead map, summary table, violin plots

    -   summer parr

Compare relative changes in capacity

# Stream Temperature Discussion
