---
title: "Upper Walla Walla Watershed - Stream Temperature Assessment"
author:
  - Bryce N. Oldemeyer:
      email: bryce.oldemeyer@mthoodenvironmental.com
      institute: [mhe_challis]
      correspondence: true
  - Mark Roes:
      email: mark.roes@mthoodenvironmental.com
      institute: [mhe_sandy]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: wdfw
      correspondence: false
  - Joshua P. Egan:
      email: joshua.egan@mthoodenvironmental.com
      institute: [mhe_moscow]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [mhe_mccall]
      correspondence: true
institute:
  - mhe_challis: Mount Hood Environmental, PO Box 1303, Challis, Idaho 83226, USA
  - mhe_sandy: Mount Hood Environmental, 39085 Pioneer Boulevard \#100 Mezzanine, Sandy, Oregon, 97055, USA
  - wdfw: Washington Department of Fish and Wildife, Fish Program, Science Division, 1111 Washington Street NE, Olympia, Washington, 98501, USA
  - mhe_moscow: Mount Hood Environmental, Temporary Moscow Address, Idaho xxxxx, USA
  - mhe_mccall: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    number_sections: yes
    pandoc_args:
    - --lua-filter=../templates/scholarly-metadata.lua
    - --lua-filter=../templates/author-info-blocks.lua
    - --lua-filter=../templates/pagebreak.lua  
csl: "../templates/journal-of-archaeological-science.csl"
bibliography:
  - AckermanLibrary.bib
always_allow_html: yes
---

<!-- the following inserts MHE logo into header -->

```{=html}
<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"logo.jpg\" style=\"float: right;width: 150px;\"/>')
   });
</script>
```
```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 600
)

options(knitr.kable.NA = '--')

```

```{r load-libraries}
# run this if running chunks directly

# for formatting
library(kableExtra)
library(ggpubr)

# for analysis
library(sf)
library(here)
library(tidyverse)
library(magrittr)
library(ggmap)
library(nngeo)
library(janitor)

theme_set(theme_pubr(x.text.angle = 45,
                     base_size = 8))

```

```{r load-data}

# Project area polygons and 200m layer (with habitat attributes) within project area

load(here("analysis/data/derived_data/uww_spatial.rda"))


```

# Background

Water is the cornerstone of First Foods production and inherently tied to the five fundamental touchstones: hydrology, geomorphology, habitat and network connectivity, riverine biotic community, and riparian vegetation [@Jones2008]. These touchstones form the Umatilla River Vision, a holistic approach to water quality restoration. Add description of stream assessment

## Study Area

This assessment will be focused on the alluvial channel of the Walla Walla River from the confluence with Dry Creek near Lowden, Washington, to the headwaters of the North and South forks of the Walla Walla River. The primary study area includes approximately 70 miles of stream. The secondary study area includes the catchment of the primary study area which encompasses approximately 885 square miles.

# Stream Temperature Assessment

The primary objectives of the Upper Walla Walla stream temperature assessment is to:

-   Summarize current stream temperatures for the Upper Walla Walla using NorWest stream temperature data (CITE)

-   Compare predictions of stream temperatures in the Upper Walla Walla for 2040 and 2080 climate scenarios.

-   Summarize temperature suitability based solely on species-specific temperature models for the Upper Walla Walla.

-   Run a hypothetical scenario were restoration efforts over the next \~60 years help decrease stream temperatures by -1^o^ C

## Stream Temperature Data

NorWest August stream temperature estimates -- 2002-2011, 2040, and 2080 summaries. Additionally, -1^o^ C temperature scenario.

-   S2_02_11 Historical composite scenario representing 10 year average August mean stream temperatures for 2002-2011

-   S30_2040D Future August mean stream temperature scenario based on global climate model ensemble average projected changes in August air temperature and stream discharge for the A1B warming trajectory in the 2040s (2030-2059). Future stream deltas within a NorWeST unit account for differential sensitivity among streams so that cold streams warm less than warm streams.

-   S32_2080D Future August mean stream temperature scenario based on global climate model ensemble average projected changes in August air temperature and stream discharge for the A1B warming trajectory in the 2080s (2070-2099). Future stream deltas within a NorWeST unit account for differential sensitivity among streams so that cold streams warm less than warm streams.

## Fish Temperature Thresholds

The focal species for the stream temeprature suitability models are:

-   Spring-run Chinook salmon (*Oncorhynchus tshawytscha*; hereafter Chinook salmon)
-   Middle Columbia River summer steelhead (*O. mykiss*; hereafter steelhead; ESA-listed threatened)
-   Columbia River bull trout (*Salvelinus confluentus*; ESA-listed threatened)
-   Pacific lamprey (*Lampetra tridentata*)

<!-- Okay, I'm going to write out my thoughts/outline for a temperature assessment. Not because I think you're off-base by any means, mostly to get them written down and available for reference... -->

<!-- -   **Summer** stream temperature assessment -->

<!--     -   Start w/ a simple temperature assessment ignoring any information on fish -->

<!--     -   This will have to be a summer stream assessment as spatially and temporally continuous datasets are unavailable to our knowledge. -->

<!--     -   Evaluate 4 scenarios: -->

<!--         -   1\. Contemporary: Based on historical scenario S2_02_11 representing 10-year average August mean stream temperatures for 2022 - 2011. -->

<!--             -   realize this is already somewhat historic, but perhaps the best contemporary dataset we have -->

<!--         -   2\. 2040 scenario -->

<!--         -   3\. 2080 scenario -->

<!--         -   4\. Contemporary minus 1: This is **scenario #1** minus 1C. I realize this is pretty unrealistic, but subtracting 1 from the 2080 scenario generally just results in something that looks like the 2040 scenario. This is just a holy grail, what if we could do something about climate change scenario. Would reducing temps from current actually make a difference? -->

<!--     -   I think the overall spatial scale could be reduced i.e., not necessarily the entire stream length available from NHDplusV2. Maybe something like: -->

<!--         -   ***Some extent used in the hydrology assessment by Rio? You might lean on Rio to define the upper extents so that it overlaps with their analysis.*** -->

<!--         -   The maximum extents among the QRF scenarios? -->

<!--         -   The maximum extents available for steelhead, Chinook, lamprey, bull trout defined in StreamNet? -->

<!--         -   Some extent defined by CTUIR as relevant to First Foods? -->

<!--     -   Because the scale is so large, you might make these 1 column of 4 maps rather than 2x2 as it currently is. -->

<!--     -   Summarized as maps, but perhaps also as box/violin plots. Spatial scale would need to be considered for summaries and perhaps use the stream partitions that will be later defined by the group. -->

<!--     -   Export these shapefiles for additional mapping and including in the report elsewhere. -->

<!-- -   Introduce temperature threshholds by species, focusing on summer life-stages\ -->

<!--     -   Chinook salmon (spawning and summer parr)\ -->

<!--     -   Steelhead (summer parr)\ -->

<!--     -   Lamprey (???) -->

<!--     -   Bull trout (???) -->

<!--     -   The lamprey and bull trout could just be more general temperature thresholds -->

<!--     -   Summarized in tables -->

<!-- -   Merge temperature scenarios with threshold information. -->

<!--     -   Would include maps similar to those included in Appendix A of the IRA that show the proportion of time that a water temperature was "within" a given life-stage specific threshold. -->

<!--     -   focus only on summer or general life-stages -->

<!--     -   also across the 4 temperature scenarios -->

<!--     -   We should think about how to summarize this in tabular form or as figures (box plots, bar plots?) form as well. -->

<!-- -   QRF assessment -->

<!--     -   Re-run capacity extrapolations, except changing the Avg August covariate to temp scenarios #2, #3, and #4. -->

<!--     -   Only needs to be done for Chinook (spawning, summer rearing) and steelhead (summer rearing) -->

<!--     -   Maybe no maps necessary? -->

<!--     -   just summarize relative changes to capacity -->

<!--     -   This is in "beta mode"...if it fails, we can fall back on all of the above -->

# Results

## Stream Temperature Results

```{r current_temp_map}

river_color = "lightskyblue1"

# Current temp map
temp_map_cur = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA)),
          aes(color = S2_02_11),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2002-2011 historical composite scenario",
       color = expression("Temperature ("^o*"C)"))

# 2040 temp map
temp_map_2040 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S30_2040D = replace(S30_2040D, S30_2040D == -9999.00, NA)),
          aes(color = S30_2040D),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2040 future scenario",
       color = expression("Temperature ("^o*"C)"))

# 2080 temp map
temp_map_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA)),
          aes(color = S32_2080D),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2080 future scenario",
       color = expression("Temperature ("^o*"C)"))

# -1C results
temp_map_1C_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA)),
          aes(color = (S32_2080D-1)),
          size = 1) +
  scale_color_viridis_c(direction = 1, limits = c(5,25)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = expression("Hypothetical restoration effect on \n 2080 future scenario (-1"^o*"C)"),
       color = expression("Temperature ("^o*"C)"))


# 2040 comparison
temp_rel_map_2040 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S30_2040D = replace(S30_2040D, S30_2040D == -9999.00, NA),
                   S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
                   temp_dif = S30_2040D - S2_02_11),
          aes(color = temp_dif),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(0,2.5)) +
#scale_fill_gradientn(limits = c(1,3)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2040 future scenario",
       color = expression("Temperature ("^o*"C)"))

# 2080 comparison
temp_rel_map_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA),
                   S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
                   temp_dif = S32_2080D - S2_02_11),
          aes(color = temp_dif),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(0,2.5)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "2080 future scenario",
       color = expression("Temperature ("^o*"C)"))


# -1C results
temp_rel_map_1C_2080 = uww_rch_sf %>%
  ggplot() +
  geom_sf(data = uww_huc_sf %>%
            st_union() %>%
            nngeo::st_remove_holes(),
          fill = NA,
          color = "gray50") +
  geom_sf(color = river_color) +
  geom_sf(data = uww_rch_sf %>%
            mutate(S32_2080D = replace(S32_2080D, S32_2080D == -9999.00, NA),
                   S2_02_11 = replace(S2_02_11, S2_02_11 == -9999.00, NA),
                   temp_dif = S32_2080D - S2_02_11),
          aes(color = (temp_dif-1)),
          size = 1) +
  scale_color_viridis_c(direction = 1,limits = c(0,2.5)) +
  #theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = expression("Hypothetical restoration effect on \n 2080 future scenario (-1"^o*"C)"),
       color = expression("Temperature ("^o*"C)"))

```

```{r current-temp-maps, fig.cap = "NorWeST average August mean stream temperatures for the Upper Walla Walla."}

ggarrange(temp_map_cur,temp_map_2040,temp_map_2080,temp_map_1C_2080)

```

```{r temp-comparison-maps, fig.cap = "Comparison of future NorWeST average August mean stream temperatures for  2040, 2080, and 2080 - 1oC to the 2002-2011 historical composite scanerio for the Upper Walla Walla."}

ggarrange(temp_map_cur,temp_rel_map_2040,temp_rel_map_2080, temp_rel_map_1C_2080)

```

## Fish Temperature Suitability Results

Chinook map & summary table

Steelhead map & summary table

Bull trout map & summary table

Lamprey map & summary table

# Stream Temperature Discussion
